Viscous Transport in Eroding Porous Media

### 2019-08-20 docs/transport/erosion_transport.tex
working on intro of transport paper

To describe the relevant hydrodynamics, we model two-dimensional
microscopic scales where the fluid is governed by the incompressible
Stokes equations (for example, see~\cite{hym-den-hag-kan2019,
den-ica-hid2018, bar-mar-vee-zha2018}).  To couple the erosion and the
hydrodynamics, the grains of the porous media are eroded at a rate that
is proportional to the hydrodynamic shear stress~\cite{wan-fel2004,
ris-moo-chi-she-zha2012, qua-moo2018}.  The incompressible Stokes
equation are solved by first recasting them as a boundary integral
equation (BIE).  The BIE is discretized with a high-order quadrature
method that resolves nearly touching bodies with a near-singular
integration scheme~\cite{bar-wu-vee2015}.  The erosion dynamics are
simulated by applying a stable second-order time stepping Runge-Kutta
method to the {\thL} formulation~\cite{hou-low-she1994} of the eroding
bodies.  Given the solution of the incompressible Stokes equations are
solved in an eroded porous geometry, we investigate different metrics to
characterize transport.  We focus on three metrics: the tortuosity, the
anomalous diffusion, and the distribution of geometrical channels.

@@

Gap sizes are important for network models ...

The numerical methods to solve the governing fluid and erosion equations
are based on previous work of the authors~\cite{qua-moo2018}.  The fluid
solver uses a second-kind integral equation that is discretized with the
trapezoid rule and solved iteratively with fast multipole accelerated
GMRES, and then the trapezoid rule is used to compute the shear stress
which is required to find the erosion rate.  The integral
equation formulation only requires discretizing the boundary of the
geometry, the complex porous matrix can be easily resolved.  Once the
erosion rate is computed, the bodies are eroded using a
{\thL}~\cite{hou-low-she1994} coordinate system.  This allows us to
maintain a discretization of the eroding bodies that is equally spaced
in arclength, and is guaranteed to not tangle.

The trapezoid rule is spectrally accurate discretization of integral
equation formulations of the two-dimensional Stokes equation.  However,
at a fixed resolution, the quadrature error of the trapezoid rule is
arbitrarily large near the eroding bodies.  Therefore, simply upsampling
to evaluate the velocity near the eroding bodies is not sufficient.  In
our previous work~\cite{qua-moo2018}, we restrict the proximity between
eroding bodies.  In addition, since we only visualized the velocity and
vorticity in the fluid bulk once per time step, we used an expensive
upsampling scheme.  However, upsampling is not practical when if it is
required at each GMRES iteration or when simulating tracers in the flow.
Therefore, in this paper, we apply a Barycentric quadrature rule to
evaluate the Stokes double-layer potential~\cite{bar-wu-vee2015}.
However, to compute derivatives of the flow, such as the vorticity and
the shear stress, we extend this work to compute the required
derivatives of the double-layer potential.

@@

We calculate the anomalous diffusion rates by analyzing the tracer
trajectories governed by equation~\eqref{eqn:trajectories} for
geometries that are formed by erosion.  Given a set of trajectories, we
define $\lambda_j(t)$ to be the arclength of the trajectory
\begin{align}
  \lambda_j(t) = \int_{0}^t \|\ss'_j(\tau)\|\, d\tau, 
    \quad j=1,\ldots,N_p.
\end{align}
Then, the first and second ensemble moments are
\begin{align}
  \langle \lambda \rangle (t) = 
    \frac{1}{N_p} \sum_{j=1}^{N_p} \lambda_j(t), \quad 
    \sigma_\lambda^2(t) = \frac{1}{N_p} 
    \left[\lambda_j(t) - \langle \lambda \rangle(t) \right]^2.
\end{align}
Then, particle dispersion is characterized as $\sqrt{\sigma_\lambda^2}$.
At early times, the particles have not explored much of the geometry,
and we expect a ballistic motion $\sigma_\lambda^2 \sim t^2$.  However,
as the particles pass the grains, their trajectories are altered, and we
expect that $\sigma_\lambda \sim t^\alpha$, with $\alpha < 1$,
indicating that the flow is sub-diffusive.  With respect to erosion, as
the bodies erode, we expect that the channelization of the geometry
results in more trajectories undergoing a ballistic motion, and this
would indicate decrease particle dispersion, meaning a larger anomalous
diffusion rate.

>>>>

To describe the relevant hydrodynamics, we model two-dimensional
microscopic scales where the fluid is governed by the incompressible
Stokes equations (for example, see~\cite{hym-den-hag-kan2019,
den-ica-hid2018, bar-mar-vee-zha2018}).  To couple the erosion and the
hydrodynamics, the grains of the porous media are eroded at a rate that
is proportional to the hydrodynamic shear stress~\cite{wan-fel2004,
ris-moo-chi-she-zha2012, qua-moo2018, par-izu2000}.  The incompressible
Stokes equation are solved by first recasting them as a boundary
integral equation (BIE).  The BIE is discretized with a high-order
quadrature method that resolves nearly touching bodies with a
near-singular integration scheme~\cite{bar-wu-vee2015}.  The erosion
dynamics are simulated by applying a stable second-order time stepping
Runge-Kutta method to the {\thL} formulation~\cite{hou-low-she1994} of
the eroding bodies.  Given the solution of the incompressible Stokes
equations are solved in an eroded porous geometry, we investigate
different metrics to characterize transport.  We focus on three metrics:
the tortuosity, the anomalous diffusion, and the distribution of
geometrical channels.

@@

Rather than solving the fluid equations in a complex porous matrix,
network models can be used to describe the flow.  Here, the pores
between neighboring grains are thought of as small channels, and the
geometry consists of a connected network of these
channels~\cite{bry-kin-mel1993, bry-mel-cad1993, bij-blu2006} with the
flow resembling a Hagen-Poiseuille profile in each channel.  So that
network models can be applied to an eroding geometry, we characterize
the pore sizes between neighboring grains as a function of the
geometry's porosity.  The pore size distribution can be used to quantify
velocity distributions~\cite{ali-par-wei-bre2017, dea-qua-bir-jua2018},
channelization~\cite{sie-ili-pri-riv-gua2019}, and
connectivity~\cite{knu-car2005, wes-blo-gra2001}.

The numerical methods to solve the governing fluid and erosion equations
are based on previous work of the authors~\cite{qua-moo2018}.  We
continue to use a boundary integral equation (BIE) formulation of the
incompressible Stokes equations and a second-order Runge-Kutta method
applied to a smoothed version of the magnitude of the shear stress.  In
our previous work, we discretized the BIE with the trapezoid rule whose
accuracy grows when evaluating layer potentials close to the fluid
boundary.  Therefore, given a fixed resolution, our previous work
required a minimum separation distance between all bodies to maintain
stability.  In recent years, many near-singular integration schemes have
been developed~\cite{oja-tor2015, kli-tor2018, hel-oja2008a,
bea-yin-wil2016, bea-lai2001, klo-bar-gre-one2013}.  In this work, we
use a Barycentric quadrature method~\cite{bar2014, bar-wu-vee2015} since
it is a non-intrusive modification of the trapezoid rule, and the error
is guaranteed to be uniformly bounded.  In this work, we extend this
quadrature method so that it can be used to compute the gradient of the
velocity which is required to compute the shear stress and the fluid
vorticity.

@@

We calculate the anomalous diffusion rates by analyzing the tracer
trajectories governed by equation~\eqref{eqn:trajectories} for
geometries that are formed by erosion.  Given a set of trajectories, we
define $\lambda_j(t)$ to be the arclength of the trajectory
\begin{align}
  \lambda_j(t) = \int_{0}^t \|\ss'_j(\tau)\|\, d\tau, 
    \quad j=1,\ldots,N_p.
\end{align}
Then, the first and second ensemble moments are
\begin{align}
  \langle \lambda \rangle (t) = 
    \frac{1}{N_p} \sum_{j=1}^{N_p} \lambda_j(t), \quad 
    \sigma_\lambda^2(t) = \frac{1}{N_p} 
    \left[\lambda_j(t) - \langle \lambda \rangle(t) \right]^2.
\end{align}
Since we compute second-order statistics, we use 1000 tracers so that
the statistics have converged~\cite{bel-sal-rin1992}.  Then, particle
dispersion is characterized as $\sqrt{\sigma_\lambda^2}$.  At early
times, the particles have not explored much of the geometry, and we
expect a ballistic motion $\sigma_\lambda^2 \sim t^2$.  However, as the
particles pass the grains, their trajectories are altered, and we expect
that $\sigma_\lambda \sim t^\alpha$, with $\alpha < 1$, indicating that
the flow is sub-diffusive.  With respect to erosion, as the bodies
erode, we expect that the channelization of the geometry results in more
trajectories undergoing a ballistic motion, and this would indicate
decrease particle dispersion, meaning a larger anomalous diffusion rate.
